*Prochlorococcus marinus* responses to light and oxygen

### 2024-03-05 Code/Manuscript_O2.Rmd
Fixed figure caption and figure references

A Generalized Additive Model (GAM) [@woodGeneralizedAdditiveModels2017] was applied to the relation of chlorophyll-specific µ, d^-1^  to photoperiod and PAR level,  for each growth [O~2~] level, and for the blue and red wavebands for growth, for each *P.marinus* strain in this study. The R package *`mgcv`* [@R-mgcv] was used to model the growth rate with smoothing terms and indicate the 90, 50 and 10% quantiles for growth rate across the levels of factors. Only growth rate estimates for which the amplitude of standard error was smaller than 30% of the fitted growth rate were included in the GAM. Our priority was the effects of ecologically relevant blue light on growth trends.  We also included GAM analyses of growth responses to red light, which is not ecophysiologically relevant, but which might prove mechanistically informative [@murphyPhotoinactivationPhotosystemII2017].

@@

CladeProchloroPhotosynthDepthO2_cap <- glue("**Ocean detection of *Prochlorococcus marinus* photosynthesis complexes.**", 
                       "Protein detections are plotted vs. O~2~ (µM) (X axis) and depth (m) (Y axis) at sample origin.",
                       "Rows separate data annotated as from *Prochlorococcus* clades: HLI (represents *P. marinus* MED4 annotated with solid black circles), LLI, LLII/III (represents *P. marinus* SS120 annotated with solid black circles) and LLIV (represents *P. marinus* MIT9313 annotated with solid black circles).",
                       "Columns show detections of proteins annotated as  Photosystem II (PSII), Cytochromeb6f complex (Cytb6f), Photosystem I (PSI), ATP Synthase or Ribulose-1,5-bisphosphate oxygenase carboxylase (RUBISCO).",
                       "For comparison culture growth experimental conditions are indicated by horizontal grey lines for depths approximating imposed peak Photosynthetically Active Radiation (µmol photons m^-2^ s^-1^); and vertical grey lines for [O~2~] (µM).",
                       "Data obtained from OceanProteinPortal (https://www.oceanproteinportal.org/).",
 .sep = " ")

@@

The representative of the LLII/III clade, *P. marinus* SS120 showed almost no growth under 2.5 µM O~2~ experiments (Figure \@ref(fig:PhotoperiodPurFitsPlots)F). Most 4 h photoperiod treatments of SS120 also did not grow under 250 µM O~2~, even when a 4 h photoperiod delivered cumulative diel PUR equivalent to other photoperiod treatments (Figure \@ref(fig:PhotoperiodPurFitsPlots)F). SS120 did not grow when exposed to more than ~1.0 x 10^6^ µmol photons m^-2^ d^-1^ of cumulative diel PUR under any spectral waveband or photoperiod combination, under 250 µM O~2~ experiments (Figure \@ref(fig:PhotoperiodPurFitsPlots)F).

Under both 25 or 250 µM O~2~ experiments, SS120 growth plateaued by about 5.0 x 10^5^ µmol photons m^-2^ d^-1^ diel PUR, with some scatter among photoperiod and spectral waveband regimes. The onset of growth inhibition extended to higher cumulative diel PUR for cultures under 25 µM O~2~, showing that SS120 is partially protected from photoinhibition of growth by 25 µM O~2~.  Under 25 µM O~2~, red light again generated more growth of SS120 per unit cumulative diel PUR, than did blue light, again consistent with lower cost of growth through lower photoinactivation under red light (Figure \@ref(fig:BluevsRedPurFitsPlots)E). For distinct fits for different photoperiods refer to figure \@ref(fig:PhotoperiodPurFitsPlots)D,E and F.

@@

The LLIV clade representative, *P. marinus* MIT9313, showed growth rising to a plateau by about 5 x 10^5^ µmol photons m^-2^ d^-1^ of cumulative diel PUR, with higher growth rates over a narrower plateau under 25 and 250 µM O~2~, compared to a wider, lower, flatter response to cumulative diel PUR under 2.5 µM O~2~ (Figure \@ref(fig:PhotoperiodPurFitsPlots)G,H). Above about 1.0 x 10^6^ µmol photons m^-2^ d^-1^ of cumulative PUR under 250 µM O~2~, MIT9313 showed full inhibition of growth, across photoperiods, and spectral wavebands (Figure \@ref(fig:PhotoperiodPurFitsPlots)G).  In contrast, under 25 µM O~2~, MIT9313 showed a greatly extended exploitation of higher cumulative diel PUR, with full growth inhibition only above about 3.5 x 10^6^ µmol photons m^-2^ d^-1^ (Figure \@ref(fig:PhotoperiodPurFitsPlots)H).  Similarly, under 2.5 µM O~2~, MIT9313 grew more slowly, but only showed full growth inhibition above about 3.5 x 10^6^ µmol photons m^-2^ d^-1^ cumulative diel PUR (Figure \@ref(fig:PhotoperiodPurFitsPlots)I).

@@

Under full atmospheric [O~2~] and blue light, LL clades of *Prochlorococcus* are restricted to growth under low light, in part because they suffer photoinhibition of Photosystem II (PSII) through several paths, including direct absorbance of UV or blue light, in parallel with generation of Reactive Oxygen Species (ROS) if the electron flow is slowed [@aroPhotoinhibitionPhotosystemII1993], producing damaging singlet oxygen (^1^O~2~) [@aroPhotoinhibitionPhotosystemII1993; @soitamoPhotoinhibitionMarinePicocyanobacteria2017; @murphyPhotoinactivationPhotosystemII2017; @hakalaEvidenceRoleOxygenevolving2005]. Repair of photoinactivated PSII relies on the removal of damaged PsbA [@mannInvolvementFtsHHomologue2000; @adamFtsHProteasesChloroplasts2005], followed by reassembly with newly synthesized PsbA [@nixonRecentAdvancesUnderstanding2010]. Degradation of PsbA is a rate-limiting step in recovery from photoinhibition [@kanervoD1ProteinDegradation1993], mediated largely by a heterohexamer of (FtsH12)~3~, a membrane-bound [@sacharzSubCellularLocation2015; @zakInitialStepsBiogenesis2001] metalloprotease  [(Chiba et al., 2002) (Yoshioka-Nishimura and Yamamoto, 2014).  [@adamFtsHProteasesChloroplasts2005; @boehmSubunitOrganizationSynechocystis2012a; @nixonRecentAdvancesUnderstanding2010; @komendaExposedNTerminalTail2007] (Pisareva et al., 2007; Sacharz et al., 2015),  @mannInvolvementFtsHHomologue2000, (AMANDA REF from ALGATECH).

*Prochlorococcus* genomes encode 4 FtsH proteins, henceforth referred to as FtsH1-4, homologs to the characterized FtsH isoforms of the model freshwater cyanobacterium *Synechocystis* FtsH, and with presumably parallel functions (Table \@ref(tab:FtsHTable)). Upon a shift to higher light HLI MED4 upregulated expression of FtsH1 and FtsH2 [@bonisteelStrainSpecificDifferences2018], homologs to the *Synechocystis* *slr*0228 and *slr*1604, implicated in PSII repair. In contrast, representative LLIV strain MIT9313 showed less overall expression of the FtsH proteases, and thus has fewer FtsH serving each photosystem.  Furthermore, MIT9313 expressed primarily FtsH3, homologous to *Synechocystis* *slr*1463, possibly involved in PSI biogenesis, and FtsH expression did not increase in response to light stress in MIT9313. Through adaptation to steady low light, clade LLIV *Prochlorococcus* instead allocate resources to processes other than dynamic regulation of PSII repair.

@@

CladeProchloroProteinMetabDepthO2_cap <- glue("**Ocean detection of *Prochlorococcus marinus* protein metabolism complexes.**", 
                       "Protein detections are plotted vs. O~2~ (µM) (X axis) and depth (m) (Y axis) at sample origin.",
                       "Rows separate data annotated as from *Prochlorococcus marinus* strains MED4 (clade HLI), SS120 (clade LLII/III) and MIT9313 (clade LLIV).",
                       "Columns show detections of proteins annotated as FtsH Protease Complexes (FtsH1, FtsH2, FtsH3) or the Ribosome).",
                       "Culture growth experimental conditions indicated by horizontal grey lines for depths approximating Photosynthetically Active Radiation (µmol photons m^-2^ s^-1^) and vertical grey lines for [O~2~] (µM).",
                       "Data obtained from OceanProteinPortal (https://www.oceanproteinportal.org/).",
 .sep = " ")

@@

West *et al*. [@westNichePartitioningProchlorococcusPopulationsStratified1999] and Malmstrom *et al*. [@malmstromTemporalDynamicsProchlorococcus2010] found that decreased abundances of the LL clades corresponded to increased depth of the surface mixed layer. Malmstrom *et al*. [@malmstromTemporalDynamicsProchlorococcus2010] attributes the transport of LL ecotypes to the surface and consequent exposure to photoinhibitory high light levels as the reason for low cell abundances with increased mixed layer depth. West *et al*.  [@westNichePartitioningProchlorococcusPopulationsStratified1999] found the depth of the mixed layer strongly influenced the depth transition from HL to LL clades, but that factors other than light levels may influence the variations in the upper and lower depth limits of these ecotypes. We hypothesize that low cell abundances of LL ecotypes in the mixed layer is likely driven in part by increased [O~2~],  and it is [O~2~] that constrains LL clades to deeper waters, not necessarily the light level. We found that under 25 µM O~2~ representatives of 'LL' clades, SS120 and MIT9313, actually tolerate approximately 1.0 x 10^6^ µmol photons m^-2^ d^-1^ of PUR (Figure \@ref(fig:PhotoperiodPurFitsPlots)B,C), comparable to the representative HL clade, MED4 which also exhibited growth saturation at the same cumulative diel PUR of 1.0 x 10^6^ µmol photons m^-2^ d^-1^ (Figure \@ref(fig:PhotoperiodPurFitsPlots)A).  Growth under lower O~2~ allowed MIT9313 to substantially increase its exploitation of higher diel PUR (Figure \@ref(fig:PhotoperiodPurFitsPlots)C).

@@

We analyzed  growth rates to determine the viability of *P. marinus* MED4, a clade HLI ecotype found near the ocean surface,
*P. marinus* SS120, a clade LLII/III ecotype found deep in the water column, and *P. marinus* MIT9313, a clade LLIV ecotype also found in deep oceans including OMZ, under a matrix of spectral, light level, photoperiod and oxygen conditions approximating present day and hypothetical future niches. Low light levels in the blue spectral waveband prevail in deep ocean waters [@morelAvailableUsableStored1978; @holtropVibrationalModesWater2021], while solar incidence angle and light attenuation also progressively attenuate the photoperiod with depth beneath the ocean surface.

*Prochlorococcus marinus* MED4 has a physiological requirement for more than 4 h of light per day; thus this strain will not exploit habitats with short photoperiods, typical of temperate winter or light attenuated depths, even if water temperature warms into the clade HLI tolerance range. MED4 is also excluded from the lowest oxygen habitats, represented by our 2.5 µM O~2~ experimental conditions. MED4 can, however, grow under OMZ regions with slightly higher [O~2~], as demonstrated by the growth under our 25 µM O~2~ experiments. Genomic analyses (Figure \@ref(fig:ProchlorococcusEnzymeKms))  and previous transcriptional analyses [@bergResponsesPsbAHli2011]  suggest MED4 is excluded from growth below ~ 25 µM O~2~ because it relies upon a ubiquinol oxidase, non-electrogenic, to maintain oxidation/reduction balance in the intersystem electron transport chain.  On the other hand, MED4 shows inducible expression of FtsH isoforms [@bonisteelStrainSpecificDifferences2018], to counter photoinactivation of PSII under higher PAR and [O~2~] environments. Photoinactivation, does, however, impose an increased cost of growth upon MED4, since growth under red light, to lower the rate of photoinactivation of PSII [@murphyPhotoinactivationPhotosystemII2017], allows MED4 to achieve faster growth per absorbed photon than growth under blue light. TARA Oceans Project data [@delmontLinkingPangenomesMetagenomes2018] indeed reported presence of *P. marinus* MED4-like genomes at depths ranging from 5 m to 90 m, representing high to low blue light levels, in the Pacific South East Ocean. Delmont and Eren [@delmontLinkingPangenomesMetagenomes2018] did not analyze data from depths beyond the subsurface chlorophyll maximum layer, the layer in the water column where the chlorophyll *a* concentration peaks, nor did they report the [O~2~] at those depths. Our growth findings are also consistent with Figure\@ref(fig:CladeProchloroPhotosynthDepthO2) showing PSII proteins annotated as MED4, clade HLI, at depths up to 200 meters with O~2~ of 15 µM.

*Prochlorococcus marinus* SS120, a LLII/III clade representative, showed an interactive inhibition of growth by oxygen and cumulative diel PUR, with a higher tolerance for higher cumulative diel PUR under 25 µM O~2~, compared to 250 µM O~2~. Thus, SS120 can exploit habitats with O~2~ levels spanning 25 µM O~2~ to 250 µM O~2~, including higher PAR environments within OMZ.  SS120 is likely excluded from the combination of higher [O~2~] and higher PAR by genomic limitations on capacity for DNA repair (Figure \@ref(fig:DNARepairFig)), and possibly by limited capacity for synthesis of reactive oxygen quenchers (Figure  \@ref(fig:ProchlorococcusEnzymeKms)).  Our growth results are supported by Lavin *et al*. [@lavinNovelLineagesProchlorococcus2010] who found evidence of LLII/III
and LLIV ecotypes, using terminal restriction fragment length polymorphism analyses, at depths above 40 m, where light levels are
higher, within OMZ, and by Figure \@ref(fig:CladeProchloroPhotosynthDepthO2) showing PSII protein subunits annotated as derived from SS120 at all depths ranging
from 20 to 200 m and all [O~2~] in an OMZ of the tropical North Pacific Ocean. SS120 grew under photoperiods longer than 4 h and showed increasing growth rate with increasing photoperiods.  However, we found *P. marinus* SS120 can potentially exploit more diverse ecological niches within the ocean layers, even in regions with higher levels of blue spectral waveband, but only under O~2~ of 25 µM (Figure \@ref(fig:SS120PAR)). Lavin *et al*. [@lavinNovelLineagesProchlorococcus2010] show evidence of LLII/III ecotypes at varying depths from 20 to 30 m and from 75 to 200 µM O~2~ in the oxygen minimum zone (OMZ) of the tropical South Pacific Ocean, illustrating its tolerance of high light levels under low O~2~. SS120 has the potential to thrive in deep temperate zones, specifically during the spring, summer, and fall seasons when the duration of daylight exceeds 4 h, if O~2~ are near surface saturation of about 250 µM. Under lower oxygen levels of 25 µM, SS120 can also potentially exploit a 4 h photoperiod in the blue waveband, and thus has the potential to inhabit a potential warmed, deep, temperate OMZ, during the winter season.

*Prochlorococcus marinus* MIT9313, a LLIV clade representative, shows potential to inhabit future warmer temperate zones year-round, as it grows under a 4 h photoperiod, expected in winter, or at light-attenuated depths. MIT9313 demonstrates an unexpected tolerance to higher light levels and cumulative diel PUR, but only under low oxygen conditions of 25 µM and 2.5 µM (Figure \@ref(fig:MIT9313PAR)), enabling MIT9313 to grow in OMZ, even at depths closer to the surface. MIT9313 carries a gene encoding (S)-2-hydroxy-acid oxidase [@cunaneCrystalStructureAnalysis2005],  with a K~M~ for [O~2~] of ~ 250 µM (Figure \@ref(fig:ProchlorococcusEnzymeKms)), which produces H~2~O~2~.  Growth at lower [O~2~] may protect MIT9313 from auto-intoxication from production of H~2~O~2~. We hypothesize that under 250 µM O~2~ and higher blue light, *P. marinus* MIT9313 suffers photoinhibition, resulting from the inactivation of PSII caused by the production of H~2~O~2~. This photoinhibition is compounded by the limited inducible repair for PSII, due to the absence of FtsH 1 and 2  expression in *P. marinus* MIT9313 [@bonisteelStrainSpecificDifferences2018].  MIT9313 shows remarkable ability to thrive under very low [O~2~], potentially allowing it to expand into broader ecological niches. These results are supported by Figure \@ref(fig:CladeProchloroPhotosynthDepthO2) showing PSI protein subunits annotated as derived from MIT9313 detected at depths \> 120 m, along with PSII subunits at depths from 50 m to 200 m in regions where O~2~ was 15 µM. The deep water environments typical for MIT 9313 are relatively nutrient rich, and *Prochlorococcus* take up and metabolize various sugars [@gomez-baenaGlucoseUptakeIts2008; @munoz-marinGlucoseUptakeProchlorococcus2017; @munoz-marinProchlorococcusCanUse2013] and amino acids [@zubkovDepthRelatedAmino2004]. In future work we aim to test whether MIT9313 is using photosynthesis to drive CO~2~ fixation in low O~2~ environments, or whether PSII generation of O~2~ acts as an electron sink for respiration, to generate using ATP for maintenance and to take up nutrients from the surroundings. Partensky *et al*. [@partenskyComparisonPhotosyntheticPerformances2018] found that in the low-light conditions found in the OMZ, MED4, SS120 and MIT9313 all became net O~2~ consumers, suggesting that the low light levels cause the respiratory chain to consume more O~2~ than the photosynthetic electron transport chain generates, thus contributing to maintenance of the low O~2~ environment.

>>>>

A Generalized Additive Model (GAM) [@woodGeneralizedAdditiveModels2017] was applied to the relation of chlorophyll-specific µ, d^-1^  to photoperiod and PAR level,  for each growth [O~2~] level, and for the blue and red wavebands for growth, for each *P. marinus* strain in this study. The R package *`mgcv`* [@R-mgcv] was used to model the growth rate with smoothing terms and indicate the 90, 50 and 10% quantiles for growth rate across the levels of factors. Only growth rate estimates for which the amplitude of standard error was smaller than 30% of the fitted growth rate were included in the GAM. Our priority was the effects of ecologically relevant blue light on growth trends.  We also included GAM analyses of growth responses to red light, which is not ecophysiologically relevant, but which might prove mechanistically informative [@murphyPhotoinactivationPhotosystemII2017].

@@

CladeProchloroPhotosynthDepthO2_cap <- glue("**Ocean detection of *Prochlorococcus marinus* photosynthesis complexes.**", 
                       "Protein detections are plotted vs. O~2~ (µM) (X axis) and depth (m) (Y axis) at sample origin.",
                       "Rows separate data annotated as from *Prochlorococcus* clades: HLI (*P. marinus* MED4 annotated with solid black circles), LLI (*P. marinus* NATL2A annotated with solid black circles), LLII/III (*P. marinus* SS120 annotated with solid black circles) and LLIV (*P. marinus* MIT9313 annotated with solid black circles).",
                       "Columns show detections of proteins annotated as  Photosystem II (PSII), Cytochromeb6f complex (Cytb6f), Photosystem I (PSI), ATP Synthase or Ribulose-1,5-bisphosphate oxygenase carboxylase (RUBISCO).",
                       "For comparison culture growth experimental conditions are indicated by horizontal grey lines for depths approximating imposed peak Photosynthetically Active Radiation (µmol photons m^-2^ s^-1^); and vertical grey lines for [O~2~] (µM).",
                       "Data obtained from OceanProteinPortal (https://www.oceanproteinportal.org/).",
 .sep = " ")

@@

The representative of the LLII/III clade, *P. marinus* SS120 showed almost no growth under 2.5 µM O~2~ experiments (Figure \@ref(fig:PhotoperiodPurFitsPlots)F). Most 4 h photoperiod treatments of SS120 also did not grow under 250 µM O~2~, even when a 4 h photoperiod delivered cumulative diel PUR equivalent to other photoperiod treatments (Figure \@ref(fig:PhotoperiodPurFitsPlots)D). SS120 did not grow when exposed to more than ~1.0 x 10^6^ µmol photons m^-2^ d^-1^ of cumulative diel PUR under any spectral waveband or photoperiod combination, under 250 µM O~2~ experiments (Figure \@ref(fig:PhotoperiodPurFitsPlots)D).

Under both 25 and 250 µM O~2~ experiments, SS120 growth plateaued by about 5.0 x 10^5^ µmol photons m^-2^ d^-1^ diel PUR, with some scatter among photoperiod and spectral waveband regimes. The onset of growth inhibition extended to higher cumulative diel PUR for cultures under 25 µM O~2~, showing that SS120 is partially protected from photoinhibition of growth by 25 µM O~2~.  Under 25 µM O~2~, red light again generated more growth of SS120 per unit cumulative diel PUR, than did blue light, again consistent with lower cost of growth through lower photoinactivation under red light (Figure \@ref(fig:BluevsRedPurFitsPlots)E). For distinct fits for different photoperiods refer to figure \@ref(fig:PhotoperiodPurFitsPlots)D,E and F.

@@

The LLIV clade representative, *P. marinus* MIT9313, showed growth rising to a plateau by about 5 x 10^5^ µmol photons m^-2^ d^-1^ of cumulative diel PUR, with higher growth rates over a narrower plateau under 25 and 250 µM O~2~ (Figure \@ref(fig:PhotoperiodPurFitsPlots)H,G), compared to a wider, lower, flatter response to cumulative diel PUR under 2.5 µM O~2~ (Figure \@ref(fig:PhotoperiodPurFitsPlots)I). Above about 1.0 x 10^6^ µmol photons m^-2^ d^-1^ of cumulative PUR under 250 µM O~2~, MIT9313 showed full inhibition of growth, across photoperiods, and spectral wavebands (Figure \@ref(fig:PhotoperiodPurFitsPlots)G).  In contrast, under 25 µM O~2~, MIT9313 showed a greatly extended exploitation of higher cumulative diel PUR, with full growth inhibition only above about 3.5 x 10^6^ µmol photons m^-2^ d^-1^ (Figure \@ref(fig:PhotoperiodPurFitsPlots)H).  Similarly, under 2.5 µM O~2~, MIT9313 grew more slowly, but only showed full growth inhibition above about 3.5 x 10^6^ µmol photons m^-2^ d^-1^ cumulative diel PUR (Figure \@ref(fig:PhotoperiodPurFitsPlots)I).

@@

Under full atmospheric [O~2~] and blue light, LL clades of *Prochlorococcus* are restricted to growth under low light, in part because they suffer photoinhibition of Photosystem II (PSII) through several paths, including direct absorbance of UV or blue light, in parallel with generation of Reactive Oxygen Species (ROS) if the electron flow is slowed [@aroPhotoinhibitionPhotosystemII1993], producing damaging singlet oxygen (^1^O~2~) [@aroPhotoinhibitionPhotosystemII1993; @soitamoPhotoinhibitionMarinePicocyanobacteria2017; @murphyPhotoinactivationPhotosystemII2017; @hakalaEvidenceRoleOxygenevolving2005]. Repair of photoinactivated PSII relies on the removal of damaged PsbA [@mannInvolvementFtsHHomologue2000; @adamFtsHProteasesChloroplasts2005], followed by reassembly with newly synthesized PsbA [@nixonRecentAdvancesUnderstanding2010]. Degradation of PsbA is a rate-limiting step in recovery from photoinhibition [@kanervoD1ProteinDegradation1993], mediated largely by a heterohexamer of (FtsH1-FtsH2)~3~, a membrane-bound [@sacharzSubCellularLocation2015; @zakInitialStepsBiogenesis2001] metalloprotease  [(Chiba et al., 2002) (Yoshioka-Nishimura and Yamamoto, 2014).  [@adamFtsHProteasesChloroplasts2005; @boehmSubunitOrganizationSynechocystis2012a; @nixonRecentAdvancesUnderstanding2010; @komendaExposedNTerminalTail2007] (Pisareva et al., 2007; Sacharz et al., 2015),  @mannInvolvementFtsHHomologue2000, (AMANDA REF from ALGATECH).

*Prochlorococcus* genomes encode 4 FtsH proteins, henceforth referred to as FtsH1-4, homologs to the characterized FtsH isoforms of the model freshwater cyanobacterium *Synechocystis* FtsH, and with presumably parallel functions (Table \@ref(tab:FtsHTable)). Upon a shift to higher light HLI MED4 upregulated expression of FtsH1 and FtsH2 [@bonisteelStrainSpecificDifferences2018], homologs to the *Synechocystis* *slr*0228 and *slr*1604, implicated in PSII repair. In contrast, representative LLIV strain MIT9313 showed less overall expression of the FtsH proteases, and thus has fewer FtsH serving each photosystem.  Furthermore, MIT9313 expressed primarily FtsH3, homologous to *Synechocystis* *slr*1463, possibly involved in PSI biogenesis, and FtsH expression did not increase in response to light stress in MIT9313 [@bonisteelStrainSpecificDifferences2018]. Through adaptation to steady low light, clade LLIV *Prochlorococcus* instead allocate resources to processes other than dynamic regulation of PSII repair.

@@

CladeProchloroProteinMetabDepthO2_cap <- glue("**Ocean detection of *Prochlorococcus marinus* protein metabolism complexes.**", 
                       "Protein detections are plotted vs. O~2~ (µM) (X axis) and depth (m) (Y axis) at sample origin.",
                       "Rows separate data annotated as from *Prochlorococcus* clades: HLI (*P. marinus* MED4 annotated with solid black circles), LLI (*P. marinus* NATL2A annotated with solid black circles), LLII/III (*P. marinus* SS120 annotated with solid black circles) and LLIV (*P. marinus* MIT9313 annotated with solid black circles).",
                       "Columns show detections of proteins annotated as FtsH Protease Complexes (FtsH1, FtsH2, FtsH3) or the Ribosome.",
                       "Culture growth experimental conditions indicated by horizontal grey lines for depths approximating Photosynthetically Active Radiation (µmol photons m^-2^ s^-1^) and vertical grey lines for [O~2~] (µM).",
                       "Data obtained from OceanProteinPortal (https://www.oceanproteinportal.org/).",
 .sep = " ")

@@

West *et al*. [@westNichePartitioningProchlorococcusPopulationsStratified1999] and Malmstrom *et al*. [@malmstromTemporalDynamicsProchlorococcus2010] found that decreased abundances of the LL clades corresponded to increased depth of the surface mixed layer. Malmstrom *et al*. [@malmstromTemporalDynamicsProchlorococcus2010] attributes the transport of LL ecotypes to the surface and consequent exposure to photoinhibitory high light levels as the reason for low cell abundances with increased mixed layer depth. West *et al*.  [@westNichePartitioningProchlorococcusPopulationsStratified1999] found the depth of the mixed layer strongly influenced the depth transition from HL to LL clades, but that factors other than light levels may influence the variations in the upper and lower depth limits of these ecotypes. We hypothesize that low cell abundances of LL ecotypes in the mixed layer is likely driven in part by increased [O~2~],  and it is [O~2~] that constrains LL clades to deeper waters, not necessarily the light level. We found that under 25 µM O~2~ representatives of 'LL' clades, SS120 and MIT9313, actually tolerate approximately 1.0 x 10^6^ µmol photons m^-2^ d^-1^ of PUR (Figure \@ref(fig:PhotoperiodPurFitsPlots)E,H), comparable to the representative HL clade, MED4 which also exhibited growth saturation at the same cumulative diel PUR of 1.0 x 10^6^ µmol photons m^-2^ d^-1^ (Figure \@ref(fig:PhotoperiodPurFitsPlots)A,B).  Growth under lower O~2~ allowed MIT9313 to substantially increase its exploitation of higher diel PUR (Figure \@ref(fig:PhotoperiodPurFitsPlots)I).

@@

We analyzed  growth rates to determine the viability of *P. marinus* MED4, a clade HLI ecotype found near the ocean surface,
*P. marinus* SS120, a clade LLII/III ecotype found deep in the water column, and *P. marinus* MIT9313, a clade LLIV ecotype also found in deep oceans including OMZ, under a matrix of spectral waveband, light level, photoperiod and oxygen conditions approximating present day and hypothetical future niches. Low light levels in the blue spectral waveband prevail in deep ocean waters [@morelAvailableUsableStored1978; @holtropVibrationalModesWater2021], while solar incidence angle and light attenuation also progressively attenuate the photoperiod with depth beneath the ocean surface.

*Prochlorococcus marinus* MED4 has a physiological requirement for more than 4 h of light per day; thus this strain will not exploit habitats with short photoperiods, typical of temperate winter or light attenuated depths, even if water temperature warms into the clade HLI tolerance range. MED4 is also excluded from the lowest oxygen habitats, represented by our 2.5 µM O~2~ experimental conditions. MED4 can, however, grow under OMZ regions with slightly higher [O~2~], as demonstrated by the growth under our 25 µM O~2~ experiments. Genomic analyses (Figure \@ref(fig:ProchlorococcusEnzymeKms))  and previous transcriptional analyses [@bergResponsesPsbAHli2011]  suggest MED4 is excluded from growth below ~ 25 µM O~2~ because it relies upon a ubiquinol oxidase, non-electrogenic, to maintain oxidation/reduction balance in the intersystem electron transport chain.  On the other hand, MED4 shows inducible expression of FtsH isoforms [@bonisteelStrainSpecificDifferences2018], to counter photoinactivation of PSII under higher PAR and [O~2~] environments. Photoinactivation, does, however, impose an increased cost of growth upon MED4, since growth under red light, to lower the rate of photoinactivation of PSII [@murphyPhotoinactivationPhotosystemII2017], allows MED4 to achieve faster growth per absorbed photon than growth under blue light. TARA Oceans Project data [@delmontLinkingPangenomesMetagenomes2018] indeed reported presence of *P. marinus* MED4-like genomes at depths ranging from 5 m to 90 m, representing high to low blue light levels, in the Pacific South East Ocean. Delmont and Eren [@delmontLinkingPangenomesMetagenomes2018] did not analyze data from depths beyond the subsurface chlorophyll maximum layer, the layer in the water column where the chlorophyll *a* concentration peaks, nor did they report the [O~2~] at those depths. Our growth findings are also consistent with Figure \@ref(fig:CladeProchloroPhotosynthDepthO2) showing PSII proteins annotated as MED4, clade HLI, at depths up to 200 meters with O~2~ of 15 µM.

*Prochlorococcus marinus* SS120, a LLII/III clade representative, showed an interactive inhibition of growth by oxygen and cumulative diel PUR, with a higher tolerance for higher cumulative diel PUR under 25 µM O~2~, compared to 250 µM O~2~. Thus, SS120 can exploit habitats with O~2~ levels spanning 25 µM O~2~ to 250 µM O~2~, including higher PAR environments within OMZ.  SS120 is likely excluded from the combination of higher [O~2~] and higher PAR by genomic limitations on capacity for DNA repair (Figure \@ref(fig:DNARepairFig)), and possibly by limited capacity for synthesis of reactive oxygen quenchers (Figure  \@ref(fig:ProchlorococcusEnzymeKms)).  Our growth results are supported by Lavin *et al*. [@lavinNovelLineagesProchlorococcus2010] who found evidence of LLII/III
and LLIV ecotypes, using terminal restriction fragment length polymorphism analyses, at depths above 40 m, where light levels are
higher, within OMZ, and by Figure \@ref(fig:CladeProchloroPhotosynthDepthO2) showing PSII protein subunits annotated as derived from SS120 at all depths ranging
from 20 to 200 m and all [O~2~] in an OMZ of the tropical North Pacific Ocean. SS120 grew under photoperiods longer than 4 h and showed increasing growth rate with increasing photoperiods.  However, we found *P. marinus* SS120 can potentially exploit more diverse ecological niches within the ocean layers, even in regions with higher levels of blue spectral waveband, but only under O~2~ of 25 µM (Figure \@ref(fig:SS120PAR)). Lavin *et al*. [@lavinNovelLineagesProchlorococcus2010] show evidence of LLII/III ecotypes at varying depths from 20 to 30 m and from 75 to 200 µM O~2~ in the oxygen minimum zone (OMZ) of the tropical South Pacific Ocean, illustrating its tolerance of high light levels under low O~2~. SS120 has the potential to thrive in deep temperate zones, specifically during the spring, summer, and fall seasons when the duration of daylight exceeds 4 h, if [O~2~] are near surface saturation of about 250 µM. Under lower oxygen levels of 25 µM, SS120 can also potentially exploit a 4 h photoperiod in the blue waveband, and thus has the potential to inhabit a potential warmed, deep, temperate OMZ, during the winter season.

*Prochlorococcus marinus* MIT9313, a LLIV clade representative, shows potential to inhabit future warmer temperate zones year-round, as it grows under a 4 h photoperiod, expected in winter, or at light-attenuated depths. MIT9313 demonstrates an unexpected tolerance to higher light levels and cumulative diel PUR, but only under low oxygen conditions of 25 µM and 2.5 µM (Figure \@ref(fig:MIT9313PAR)), enabling MIT9313 to grow in OMZ, even at depths closer to the surface. MIT9313 carries a gene encoding (S)-2-hydroxy-acid oxidase [@cunaneCrystalStructureAnalysis2005],  with a K~M~ for [O~2~] of ~ 250 µM (Figure \@ref(fig:ProchlorococcusEnzymeKms)), which produces H~2~O~2~.  Growth at lower [O~2~] may protect MIT9313 from auto-intoxication from production of H~2~O~2~. We hypothesize that under 250 µM O~2~ and higher blue light, *P. marinus* MIT9313 suffers photoinhibition, resulting from the inactivation of PSII caused by the production of H~2~O~2~. This photoinhibition is compounded by the limited inducible repair for PSII, due to the absence of FtsH 1 and 2  expression in *P. marinus* MIT9313 [@bonisteelStrainSpecificDifferences2018].  MIT9313 shows remarkable ability to thrive under very low [O~2~], potentially allowing it to expand into broader ecological niches. These results are supported by Figure \@ref(fig:CladeProchloroPhotosynthDepthO2) showing PSI protein subunits annotated as derived from MIT9313 detected at depths \> 120 m, along with PSII subunits at depths from 50 m to 200 m in regions where O~2~ was 15 µM. The deep water environments typical for MIT9313 are relatively nutrient rich, and *Prochlorococcus* take up and metabolize various sugars [@gomez-baenaGlucoseUptakeIts2008; @munoz-marinGlucoseUptakeProchlorococcus2017; @munoz-marinProchlorococcusCanUse2013] and amino acids [@zubkovDepthRelatedAmino2004]. In future work we aim to test whether MIT9313 is using photosynthesis to drive CO~2~ fixation in low O~2~ environments, or whether PSII generation of O~2~ acts as an electron sink for respiration, using ATP for maintenance and to take up nutrients from the surroundings. Partensky *et al*. [@partenskyComparisonPhotosyntheticPerformances2018] found that in the low-light conditions found in the OMZ, MED4, SS120 and MIT9313 all became net O~2~ consumers, suggesting that the low light levels cause the respiratory chain to consume more O~2~ than the photosynthetic electron transport chain generates, thus contributing to maintenance of the low O~2~ environment.
